Online methods

### 2013-06-19 docs/manuscript/methods.tex
updating methods

We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions. We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}. The idea behind the method is that variants related to a particular phenotype may affect cell-type-specific gene expression by changing regulatory elements in cell types relevant to that phenotype. Hence, overlap between associated variants and chromatin marks should occur preferentially in the relevant cell type(s). Our analysis tested for enrichment of these chromatin marks in 34 different cell types from 4 different broad tissue types.

@@

Cell-type-specificity scores per locus were estimated by normalizing BestSNP scores so that the sum of scores for a given locus across all cell types equaled 1. Cell-type-specificity scores per tissue were defined by summing normalized BestSNP scores across all loci within a given tissue. 10,000 sets of SNPs (matched to the IndexSNP regions having the same total number of H3K4me3 peaks) were sampled (from among background SNPs provided with the software) to estimate null distributions of cell-type-specificity scores per locus and per tissue. P-values for cell-type-specificity scores summed across all BestSNPs for each tissue (the observed per-tissue score) were estimated as the proportion of random SNP sets with a per-tissue score exceeding the observed per- tissue score. We identified loci with BestSNP cell-type-specificity scores falling at or above the 95th percentile of the corresponding null distribution as demonstrating greater than expected specificity within a given cell type (50).
Figure S20 shows p-values for the cell-type-specific overlap of H3K4me3 marks and IndexSNP regions for each cell type. Four cell/tissue types showed significant overlap at nominal p ≤ 0.05: for EduYears, anterior caudate (p = 0.0089), CD4+ naive primary cells (p = 0.032), hippocampus middle (p = 0.05) and muscle satellite cultured cells (p = 0.0236); and for College, anterior caudate (p = 0.0007). Additionally, for College the mid-frontal lobe showed marginal enrichment at p = 0.0502. Only the anterior caudate results for College survive correction for multiple hypothesis testing.
The results from the analysis of overlap between H3K4me4 chromatin marks and education-associated SNP regions suggest gene expression regulatory function for those loci in specific cell types. In particular, we note that anterior caudate tissue appears enriched for both EduYears and College phenotypes, although only the latter survives multiple testing correction. Figure S21 shows cell-type-specificity scores per locus in the four nominally significant tissues and 95th-percentile threshold (dashed red line). Loci above the threshold were identified as showing greater than chance specificity within that particular cell type. Table S19 identifies these enriched loci, along with distance to the nearest chromatin mark.

@@

I like when chromosomes interact because I like epistasis. Also there is a cool figure with lots of circles.

@@

>>>>

We employed a recently published method (http://www.broadinstitute.org/mpg/epigwas/) \cite{Trynka2013, Rietvald2013} that tests for cell-type-specific enrichment of active chromatin, measured through H3K4me3 chromatin marks \cite{Koch2007} in regions surrounding the 731 SNPs that comprise the 501 discovery interactions. We subsequently refer to these 731 SNPs as the IndexSNPs A recent paper shows that H3K4me3 chromatin marks are the most cell-type-specific marks in terms of co-localization with previously published GWAS loci \cite{Trynka2013}. The idea behind the method is that variants related to a particular phenotype may affect cell-type-specific gene expression by changing regulatory elements in cell types relevant to that phenotype. Hence, overlap between associated variants and chromatin marks should occur preferentially in the relevant cell type(s). Our analysis tested for enrichment of these chromatin marks in 34 different cell types from 4 different broad tissue types.

@@

Cell-type-specificity scores per locus were estimated by normalizing BestSNP scores so that the sum of scores for a given locus across all cell types equaled 1. Cell-type-specificity scores per tissue were defined by summing normalized BestSNP scores across all loci within a given tissue. 10,000 sets of SNPs (matched to the IndexSNP regions having the same total number of H3K4me3 peaks) were sampled (from among background SNPs provided with the software) to estimate null distributions of cell-type-specificity scores per locus and per tissue. P-values for cell-type-specificity scores summed across all BestSNPs for each tissue (the observed per-tissue score) were estimated as the proportion of random SNP sets with a per-tissue score exceeding the observed per-tissue score. We identified loci with BestSNP cell-type-specificity scores falling at or above the 95th percentile of the corresponding null distribution as demonstrating greater than expected specificity within a given cell type \cite{Trynka2013}.

@@

It has been shown \cite{Lieberman-Aiden2009} that different regions on different chromosomes or within chromosomes spatially colocalise within the cell. We shall refer to the colocalisation of two chromosome regions as a chromosome interaction. A map of all known pairwise chromosome interactions for K562 blood cell lines was recently produced \cite{Lan2012}, and we hypothesised that part of the underlying biological mechanism behind some of the 501 epistatic interactions may arise from chromosome interactions. We found that 44 of the putative epistatic interactions were amongst SNPs that were within 5Mb of known chromosome interactions. This means that SNP A was no more than 2.5Mb from the focal point of the chromosome interaction on chromosome A, and SNP B was no more than 2.5Mb from the focal point on chromosome B.

We performed simulations to test how extreme the observation of 44 epistatic interactions overlapping with chromosome interactions is compared to chance. Chromosome interactions fall within functional genomic regions \cite{Lieberman-Aiden2009, Lan2012}, and the SNPs in our epistatic interactions are enriched for functional genomic regions. Therefore, we designed the simulations to ensure that the null distribution was of chromosome interactions between SNPs enriched for functional genomic regions but with no known epistatic interactions. To do this we used the 731 SNPs that form the 501 putative epistatic interactions and randomly shuffled them to create new sets of 501 pairs, disallowing any SNP combinations that were in the original set. Therefore, each new random set was enriched for functional regions but had no genetic interactions. We scanned the map of chromosome interactions for overlaps with the new sets and then repeated the random shuffling process. We performed 10,000 such permutations to generate a null distribution of chromosome interaction overlaps.

We repeated this process, searching for overlaps within 1Mb, 250kb, and 10kb.

@@

We tested for enrichment of genic features for the 731 SNPs that comprise the 501 epistatic interactions. For each of the 731 SNPs we calculated LD with all neighbouring SNPs within a radius of 1kb and kept all neighbouring SNPs with LD $r^2 > 0.8$. We then cross-referenced the remaining regional SNPs with the annotated reference genome (version HG19) querying whether the regional SNPs fell in intergenic, promotor, intronic, exonic, splice sites, 5' UTR or 3' UTR positions. Therefore a particular query SNP might have regional SNPs in multiple genic features.

We then performed the whole querying process for each of the 528,509 SNPs present in the SNP chip used in the scan, and used the results from this second analysis to establish the expected proportion of SNPs for each genic feature.
